\vspace{-2.5em} Spatio-angular transfer functions for fluorescence microscopes\vspace{-1em}

### 2018-04-23 notes/2018-02-23-spatio-angular-kernel/report/report.tex
Change to object scale coordinates; Patrick's Apr 20 edits.

We use plain roman type for scalars, e.g., $x, y, z$; bold lowercase roman type
for two-dimensional vectors, e.g., $\mb{r}$; hats for unit vectors, e.g.,
$\mb{\hat{s}}$; and bold capital roman type for matrices, e.g.,
$\mb{R}$. We use the real spherical harmonic functions
\begin{align}
  y_l^m(\vartheta, \varphi) =
  \begin{cases}
    \sqrt{2}K_l^m\cos(m\varphi)P_l^m(\cos\vartheta), & m > 0\\
    K_l^0P_l^0(\cos\vartheta), & m = 0\\
    \sqrt{2}K_l^m\sin(-m\varphi)P_l^{-m}(\cos\vartheta), & m < 0\\
  \end{cases}
\end{align}
where
\begin{align}
  K_l^m = \sqrt{\frac{(2l+1)}{4\pi}\frac{(l-|m|)!}{(l+|m|)!}},
\end{align}
and $P_l^m(x)$ are the associated Legendre polynomials. The $l=0$ and $l=1$
spherical harmonics are given by
\begin{equation}
\begin{aligned}
  y_0^0(\vartheta, \varphi) &= \sqrt{\frac{1}{4\pi}},\\
  y_1^{-1}(\vartheta, \varphi) = \sqrt{\frac{3}{4\pi}}\sin\varphi\sin\vartheta, \hspace{2em} y_1^0(\vartheta, \varphi) &= \sqrt{\frac{3}{4\pi}}\cos\vartheta, \hspace{2em} y_1^1(\vartheta, \varphi) = \sqrt{\frac{3}{4\pi}}\cos\varphi\sin\vartheta. \label{eq:harmonics}
\end{aligned}
\end{equation}

@@

Figure \ref{fig:frames_a} shows a schematic of the fluorescence microscope that
we are considering with a summary of our notation. We start by following Backer
and Moerner \cite{backer2014} to find the electric field at position $\rb{}$ in
the back focal plane due to a single dipole emitter at position $\ro{}$ oriented
along $\so{}$ as
\begin{align}
  \mb{\tilde{e}}_b(\rb{};\ro{}, \so{}) \propto \me^{-i(kn_o/f_o)\rb{}\cdot\ro{}}\sqrt{\frac{1}{\rho_b}}
  \begin{bmatrix}
    \sin^2\phi_b + \rho_b\cos^2\phi_b&\sin\phi_b\cos\phi_b(\rho_b - 1)&-\frac{r_b}{f_o}\cos\phi_b\\
    \sin\phi_b\cos\phi_b(\rho_b - 1)&\cos^2\phi_b + \rho_b\sin^2\phi_b&-\frac{r_b}{f_o}\sin\phi_b\\
    0&0&0
  \end{bmatrix}
  \begin{bmatrix}
    \cos\varphi_o\sin\vartheta_o\\
    \sin\varphi_o\sin\vartheta_o\\
    \cos\vartheta_o
  \end{bmatrix}
\Pi\left(\frac{r_b}{r_b^{\text{max}}}\right)\label{eq:bfp},
\end{align}
where we define $\rho_b \equiv \sqrt{1 - \left(\frac{r_b}{f_o}\right)^2}$, and
$\Pi(x)$ is a boxcar function that returns 1 when $|x| < 1$ and 0 otherwise. We
can understand this expression term by term---the exponential term accounts for
the phase as dipole emitter is moved in object space, the square root term
conserves power before and after the objective lens, the matrix models the
dipole emission process and electric field rotation caused by the objective
lens, the vector is the dipole orientation unit vector, and
$\Pi\left(\frac{r_b}{r_b^{\text{max}}}\right)$ accounts for the numerical
aperture of the lens with $r_b^{\text{max}} = \frac{f_0}{n_0}\text{NA}$. We use
a tilde to mark single dipole response functions---we will consider the response
due to a field of dipoles in the next section.

@@

If the tube lens is weakly focusing ($f_o \ll f_t$) then we can find the
electric field in the detector plane by taking the Fourier transform of the
electric field in the back focal plane
\begin{align}
  \mb{\tilde{e}}_d(\rd{}; \ro{}, \so{}) \propto \int_{\mathbb{R}^2} d\rb{}\, \tilde{\mb{e}}_b(\rb{}; \ro{}, \so{})\me^{-i (kn_t/f_t) \rb{} \cdot \rd{}}. \label{eq:ed}
\end{align}
By isolating the phase term in Eq. \ref{eq:bfp} with
$\mb{\tilde{e}}_b(\rb{};\ro{}, \so{}) \equiv \me^{-i(kn_o/f_o)\rb{}\cdot\ro{}}
\tilde{\underline{\mb{e}}}_b(\rb{};\so{})$, and plugging into Eq. \ref{eq:ed} we find
that 
\begin{align}
  \mb{\tilde{e}}_d(\rd{} - M\ro{}, \so{}) \propto \int_{\mathbb{R}^2} d\rb{}\, \tilde{\underline{\mb{e}}}_b(\rb{}; \so{})\me^{-i (kn_t/f_t)\, \rb{} \cdot [\,\rd{} - M\ro{}]}.
\end{align}
where $M = -\frac{n_o}{n_t}\frac{f_t}{f_o}$ is the transverse magnification.  By
writing the electric field in the detector plane in terms of $\rd{} - M\ro{}$,
we have established that the electric field in the detector plane is
\textit{transverse shift-invariant}---a transverse shift of the object creates a
magnified transverse shift of the image. We define coordinates on the detector
centered on the image of the object
$\rd{}' \equiv \rd{} - M\ro{} = r_d'\cos\phi_d'\mh{x} + r_d'\sin\phi_d'\mh{y}$,
then we follow Novotny \cite{nov2006} by writing the integrals in polar
coordinates, evaluating the azimuthal integrals, and writing the result
concisely in terms of three radial integrals
\begin{align}
  \tilde{\mb{e}}_d(\rd{}', \so{}) &\propto
  \begin{bmatrix}
    a(r_d') + c(r_d')\cos(2\phi_d') & c(r_d')\sin(2\phi_d') & 2ib(r_d')\cos\phi_d'\\
    c(r_d')\sin(2\phi_d') & a(r_d') - c(r_d')\cos(2\phi_d') & 2ib(r_d')\sin\phi_d'\\
    0&0&0\\
  \end{bmatrix}
  \begin{bmatrix}
    \cos\varphi_o\sin\vartheta_o\\
    \sin\varphi_o\sin\vartheta_o\\
    \cos\vartheta_o
  \end{bmatrix},\label{eq:elec}
\end{align}
where $a(r_d'), b(r_d')$, and $c(r_d')$ are
\begin{align}
  a(r_d') &= \int_0^{\theta_{\text{max}}}d\theta\ \sqrt{\cos\theta}\sin\theta(1+\cos\theta)J_0(k_b r_d'\sin\theta f_o/f_t),\\%\ \text{exp}\left\{ik_bz_o[1-(1/2)(f_o/f_t)^2\sin^2\theta]\right\},\\
  b(r_d') &= \int_0^{\theta_{\text{max}}}d\theta\ \sqrt{\cos\theta}\sin^2\theta J_1(k_b r_d'\sin\theta f_o/f_t),\\%\ \text{exp}\left\{ik_bz_o[1-(1/2)(f_o/f_t)^2\sin^2\theta]\right\},\\
  c(r_d') &= \int_0^{\theta_{\text{max}}}d\theta\ \sqrt{\cos\theta}\sin\theta(1-\cos\theta)J_2(k_b r_d'\sin\theta f_o/f_t).%\ \text{exp}\left\{ik_bz_o[1-(1/2)(f_o/f_t)^2\sin^2\theta]\right\}.
\end{align}
We can identify Eq. \ref{eq:elec} as the vector-valued \textit{coherent spread
  function} (CSF) of the microscope. The more familiar scalar-valued
CSF---sometimes called the amplitude transfer function---is only applicable to
cases where electromagnetic optics plays no role in the microscope---not true
for samples that contain dipole emitters.

To build our intuition about the CSF, we will rewrite the matrix multiplication
in Eq. \ref{eq:elec} in terms of the spherical harmonics. Notice that the $l=1$
spherical harmonics are the Cartesian components of the unit dipole axis (see
Eq. \ref{eq:harmonics}), so the CSF becomes
\begin{align}
  \tilde{\mb{e}}_d(\rd{}', \so{}) &\propto
  \begin{bmatrix}
    [a(r_d') + c(r_d')\cos(2\phi_d')]y_1^{1}(\so{}) + c(r_d')\sin(2\phi_d')y_1^{-1}(\so{}) + 2ib(r_d')\cos\phi_d'y_1^{0}(\so{})\\
    c(r_d')\sin(2\phi_d')y_1^{1}(\so{}) + [a(r_d') - c(r_d')\cos(2\phi_d')]y_1^{-1}(\so{}) + 2ib(r_d')\sin\phi_d'y_1^{0}(\so{})\\
    0\\
  \end{bmatrix}.
\end{align}
By applying the paraxial approximation ($\sin\theta\approx\theta$ and
$\cos\theta\approx 1$), the integrals $a(r_d'), b(r_d')$ and $c(r_d')$ can be
evaluated in terms of Bessel functions. We can evaluate $a(r_d')$ and $b(r_d')$ with the
help of $\int_0^zdx\ xJ_0(ax) = zJ_1(az)/a$ and
$\int_0^zdx\ x^2J_1(ax) = z^2J_2(az)/a$, respectively, and $c(r_d') = 0$ because
$J_2(x)$ is zero to first order. In this case, the CSF simplifies to
\begin{align}
  \tilde{\mb{e}}^{(p)}_d(\rd{}', \so{}) &\propto
  \begin{bmatrix}
    a^{(p)}(r_d')y_1^{1}(\so{}) + 2ib^{(p)}(r_d')\cos\phi_d'y_1^{0}(\so{})\\
    a^{(p)}(r_d')y_1^{-1}(\so{}) + 2ib^{(p)}(r_d')\sin\phi_d'y_1^{0}(\so{})\\
    0\\
  \end{bmatrix}\label{eq:paracsf},
\end{align}
where the integrals evaluate to
\begin{align}
  {a^{(p)}}(r_d') = \frac{J_1(2\pi \nu_or_d')}{\pi \nu_or_d'}, 
  &\hspace{2em}
    {b^{(p)}}(r_d') = \frac{\text{NA}}{n_o}\left[\frac{J_2(2\pi \nu_or_d')}{\pi \nu_or_d'}\right],  \label{eq:abparadef}
  \intertext{and we have substituted}
  \nu_o \equiv \frac{\text{NA}}{M\lambda},&\hspace{2em}
  \text{NA} \equiv n_o\sin\theta_{\text{max}}.
\end{align}
Under the paraxial approximation the electric fields on the detector created by
a single dipole are composed of two parts, a parallel part and a radial
part. The coefficients of the $y_1^0$ spherical harmonic in Eq. \ref{eq:paracsf}
create the radial part of the field---the $z$ component of the dipole generates
a radial field on the detector with respect to the image point. The coefficients
of the $y_1^1$ and $y_1^{-1}$ spherical harmonics create a parallel field in the
detector plane parallel to the $x$ and $y$ components of the
dipole. As the NA increases and the paraxial approximation no longer applies, we
begin to see coupling between the $x$ and $y$ components of the dipole with the $y$ and $x$ transverse fields, respectively.

We can find the intensity in the detector plane due to a single dipole
by taking the squared modulus of the CSF
\begin{align}
  h(\rd{}', \so{}) = \left|\tilde{\mb{e}}_d(\rd{}', \so{})\right|^2 \label{eq:kernel}.
\end{align}
For convenience we keep the CSF written in terms of the spherical harmonics and
use a table of spherical harmonic products (see Appendix \ref{realspherical}) to
calculate the intensity in the detector plane as
\begin{equation}
  \begin{split}
  h(\rd{}', \so{}) \propto &\left(a^2(r_d') + 2b^2(r_d') + c^2(r_d')\right)y_0^0(\so{}) -\frac{2\sqrt{15}}{5}a(r_d')c(r_d')\sin(2\phi_d')y_2^{-2}(\so{})\\ &+ \frac{1}{\sqrt{5}}\left(-a^2(r_d') + 4b^2(r_d') - c^2(r_d')\right)y_2^{0}(\so{}) +\frac{2\sqrt{15}}{5}a(r_d')c(r_d')\cos(2\phi_d')y_2^{2}(\so{}). \label{eq:kerna}
\end{split}
\end{equation}
We can identify Eq. \ref{eq:kerna} as the spatio-angular \textit{point spread
  function} (PSF) of the microscope. Writing the spatio-angular PSF in terms of
spherical harmonic functions has two advantages. First, it allows us to express
the spatio-angular PSF very concisely. Instead of considering the point spread
function for every possible dipole orientation, we only need to consider four
spatio-angular PSFs---one for each spherical harmonic. Second, the spherical
harmonic functions form an orthonormal basis for functions on the sphere---a
convenient fact that we will use later.

@@

The spatio-angular PSF under the paraxial approximation is given by
\begin{align}
      h^{(p)}(\rd{}', \so{}) \propto \left({a^{(p)}}^2(r_d') + 2{b^{(p)}}^2(r_d')\right)y_0^0(\so{}) + \frac{1}{\sqrt{5}}\left(-{a^{(p)}}^2(r_d') + {4b^{(p)}}^2(r_d')\right)y_2^{0}(\so{})\label{eq:para}.
\end{align}
We rewrite the PSF in the following form to make plotting and interpretation easier
\begin{align}
  h^{(p)}(\rd{}', \so{}) &= {h_0^0}^{(p)}(\rd{}')y_0^0(\so{}) + {h_2^0}^{(p)}(\rd{}')y_2^0(\so{}),\\
  {h_0^0}^{(p)}(\rd{}') &\equiv {a^{(p)}}^2(r_d') + 2{b^{(p)}}^2(r_d'),\\
  {h_2^0}^{(p)}(\rd{}') &\equiv \frac{1}{\sqrt{5}}\left[- {a^{(p)}}^2(r_d') + 4{b^{(p)}}^2(r_d')\right]. \label{eq:concisepsf}
\end{align}

First, consider the coefficient on the $y_0^0$ spherical harmonic in Eq.
\ref{eq:para}. This coefficient is the point spread function for an angularly
uniform distribution of fluorophores. The first term of the coefficient
${A^{(p)}}^2(r_d')$ is the familiar Airy disk that arises from the contribution
of dipoles oriented in the transverse plane, while the second term
${B^{(p)}}^2(r_d')$ is a smaller factor that arises from dipoles oriented
outside of the transverse plane. This leads to an interesting conclusion---a
uniform distribution of dipoles has a point spread function that is slightly
wider than an Airy disk even in the paraxial approximation. The Airy disk is
usually derived using paraxial scalar optics while here we have used paraxial
electromagnetic optics. Therefore, we can consider the second term to be an
electromagnetic correction to the Airy disk. We will quantify this difference in
the next section.

Next, consider the coefficient on the $y_2^0$ spherical harmonic in
Eq. \ref{eq:para}. This coefficient is the spatial PSF for a distribution of
fluorophores proportional to $3\cos^2\vartheta_o - 1$. Counterintuitively, this
fluorophore distribution cannot exist because it would require a negative number
of fluorophores along some orientations; but if this distribution could exist,
then this coefficient would be its spatial PSF. Considering negative
distributions of fluorophores in our calculations should not be cause for
concern. The spherical harmonics span the space of functions on the sphere, so
any positive fluorophore distribution can be represented by the spherical
harmonics and we never need to consider negative fluorophores.

@@

We can find the electric field on the detector created by this object by
multiplying $\bs{\mu}(\ro{}, \so{})$ with the CSF then integrating over all
positions and orientations in the object
\begin{align}
  \mb{e}_d(\rd{}) = \int_{\mathbb{S}^2}d\so{}\int_{\mathbb{R}^2}d\ro{} \ \tilde{\mb{e}}_d(\rd{} - M\mb{r}_o, \so{})\bs{\mu}(\ro{}, \so{}).\label{eq:eforward}
\end{align}
Note that these integrals represent a vector sum---the coherence of the electric
fields radiated by the object can cause cancellations of the fields created on
the detector.

>>>>

We use plain roman type for scalars, e.g., $x, y, z$; bold lowercase roman type
for two-dimensional vectors, e.g., $\mb{r}$; hats for unit vectors, e.g.,
$\mb{\hat{s}}$; and bold capital roman type for matrices, e.g.,
$\mb{R}$. We use the real spherical harmonic functions
\begin{align}
  y_l^m(\vartheta, \varphi) =
  \begin{cases}
    \sqrt{2}K_l^m\cos(m\varphi)P_l^m(\cos\vartheta), & m > 0\\
    K_l^0P_l^0(\cos\vartheta), & m = 0\\
    \sqrt{2}K_l^m\sin(-m\varphi)P_l^{-m}(\cos\vartheta), & m < 0\\
  \end{cases}
\end{align}
where
\begin{align}
  K_l^m = \sqrt{\frac{(2l+1)}{4\pi}\frac{(l-|m|)!}{(l+|m|)!}},
\end{align}
and $P_l^m(x)$ are the associated Legendre polynomials. The $l=0, 1, 2$
spherical harmonics are given by
\begin{align}
  y_0^0(\vartheta, \varphi) &= \sqrt{\frac{1}{4\pi}}, \nonumber \\ 
  y_1^{-1}(\vartheta, \varphi) = \sqrt{\frac{3}{4\pi}}\sin\varphi\sin\vartheta, \hspace{2em} y_1^0(\vartheta, \varphi) &= \sqrt{\frac{3}{4\pi}}\cos\vartheta, \hspace{2em} y_1^1(\vartheta, \varphi) = \sqrt{\frac{3}{4\pi}}\cos\varphi\sin\vartheta, \nonumber \\
  y_2^{-2}(\vartheta, \varphi) = \sqrt{\frac{15}{4\pi}}\sin\varphi\cos\varphi\sin^2\vartheta, &\hspace{2em} y_2^{-1}(\vartheta, \varphi) = \sqrt{\frac{15}{4\pi}}\sin\varphi\sin\vartheta\cos\vartheta,\nonumber \\
  y_2^0(\vartheta, \varphi) &= \sqrt{\frac{5}{16\pi}}(3\cos^2\vartheta - 1), \nonumber \\ 
  y_2^{1}(\vartheta, \varphi) = \sqrt{\frac{15}{4\pi}}\cos\varphi\sin\vartheta\cos\vartheta, &\hspace{2em} y_2^{2}(\vartheta, \varphi) = \sqrt{\frac{15}{4\pi}}(\cos^2\varphi - \sin^2\varphi)\sin^2\vartheta.\label{eq:harmonics}
\end{align}

@@

Figure \ref{fig:frames_a} shows a schematic of the fluorescence microscope that
we are considering with a summary of our notation. We start by following Backer
and Moerner \cite{backer2014} to find the electric field at position $\rb{}$ in
the back focal plane due to a single dipole emitter at position $\ro{}$ oriented
along $\so{}$ as
\begin{align}
  \mb{\tilde{e}}_b(\rb{};\ro{}, \so{}) \propto\, &\me^{-i(kn_o/f_o)\rb{}\cdot\ro{}}\sqrt{\frac{1}{\rho_b}} \times \nonumber \\
  &\begin{bmatrix}
    \sin^2\phi_b + \rho_b\cos^2\phi_b&\sin\phi_b\cos\phi_b(\rho_b - 1)&-\frac{r_b}{f_o}\cos\phi_b\\
    \sin\phi_b\cos\phi_b(\rho_b - 1)&\cos^2\phi_b + \rho_b\sin^2\phi_b&-\frac{r_b}{f_o}\sin\phi_b\\
    0&0&0
  \end{bmatrix}
  \begin{bmatrix}
    \cos\varphi_o\sin\vartheta_o\\
    \sin\varphi_o\sin\vartheta_o\\
    \cos\vartheta_o
  \end{bmatrix}
\Pi\left(\frac{r_b}{r_b^{\text{max}}}\right)\label{eq:bfp},
\end{align}
where we define $\rho_b \equiv \sqrt{1 - \left(\frac{r_b}{f_o}\right)^2}$, and
$\Pi(x)$ is a boxcar function that returns 1 when $|x| < 1$ and 0 otherwise. We
can understand this expression term by term---the exponential term accounts for
the phase as the dipole emitter is moved in object space, the square root term
conserves power before and after the objective lens, the matrix models the
dipole emission process and electric field rotation caused by the objective
lens, the vector is the dipole orientation unit vector, and
$\Pi\left(\frac{r_b}{r_b^{\text{max}}}\right)$ accounts for the numerical
aperture of the lens with $r_b^{\text{max}} = \frac{f_0}{n_0}\text{NA}$. We use
a tilde to mark single dipole response functions---we will consider the response
due to a field of dipoles in the next section.

@@

If the tube lens is weakly focusing ($f_o \ll f_t$) then we can find the
electric field in the detector plane by taking the Fourier transform of the
electric field in the back focal plane
\begin{align}
  \mb{\tilde{e}}_d(\rd{}; \ro{}, \so{}) \propto \int_{\mathbb{R}^2} d\rb{}\, \tilde{\mb{e}}_b(\rb{}; \ro{}, \so{})\me^{-i (kn_t/f_t) \rb{} \cdot \rd{}}. \label{eq:ed}
\end{align}
By isolating the phase term in Eq. \ref{eq:bfp} with
$\mb{\tilde{e}}_b(\rb{};\ro{}, \so{}) \equiv \me^{-i(kn_o/f_o)\rb{}\cdot\ro{}}
\tilde{\underline{\mb{e}}}_b(\rb{};\so{})$, and plugging into Eq. \ref{eq:ed} we find
that 
\begin{align}
  \mb{\tilde{e}}_d(\rd{} - M\ro{}, \so{}) \propto \int_{\mathbb{R}^2} d\rb{}\, \tilde{\underline{\mb{e}}}_b(\rb{}; \so{})\me^{-i (kn_t/f_t)\, \rb{} \cdot [\,\rd{} - M\ro{}]}. 
\end{align}
where $M = -\frac{n_o}{n_t}\frac{f_t}{f_o}$ is the transverse magnification.

By writing the electric field in the detector plane in terms of
$\rd{} - M\ro{}$, we have established that the imaging system magnifies and
applies a shift-invariant blur to the electric field. We can rewrite this system
as a linear shift invariant system by following Barrett section 7.2.7
\cite{barrett2004}. We scale the image
$\tilde{\mb{e}}_d'(\rd{}, \so{}) \equiv \tilde{\mb{e}}_d(M\rd{}, \so{})$ and its
coordinates $\rd{}' \equiv \rd{}/M$ so that all length scales are measured in
object space. With these definitions we can write the electric field on the
detector as $\tilde{\mb{e}}_d'(\rd{}' - \ro{})$ which is a linear shift
invariant point response function. We will continue to use a prime to denote
quantities that have been demagnified into object space.

We define a convenient set of coordinates centered on the object and scaled to
object space
$\ro{}' \equiv \rd{}' - \ro{} = r_o'\cos\phi_o'\mh{x} + r_o'\sin\phi_o'\mh{y}$.
Next, we follow Novotny \cite{nov2006} by writing the integrals in polar
coordinates, evaluating the azimuthal integrals, and writing the result
concisely in terms of three radial integrals
\begin{align}
  \tilde{\mb{e}}'_d(\ro{}', \so{}) &\propto
  \begin{bmatrix}
    a(r_o') + c(r_o')\cos(2\phi_o') & c(r_o')\sin(2\phi_o') & 2ib(r_o')\cos\phi_o'\\
    c(r_o')\sin(2\phi_o') & a(r_o') - c(r_o')\cos(2\phi_o') & 2ib(r_o')\sin\phi_o'\\
    0&0&0\\
  \end{bmatrix}
  \begin{bmatrix}
    \cos\varphi_o\sin\vartheta_o\\
    \sin\varphi_o\sin\vartheta_o\\
    \cos\vartheta_o
  \end{bmatrix},\label{eq:elec}
\end{align}
where $a(r_o'), b(r_o')$, and $c(r_o')$ are
\begin{align}
  a(r_o') &= \int_0^{\theta_{\text{max}}}d\theta\ \sqrt{\cos\theta}\sin\theta(1+\cos\theta)J_0(k_b r_o'\sin\theta f_o/f_t),\\%\ \text{exp}\left\{ik_bz_o[1-(1/2)(f_o/f_t)^2\sin^2\theta]\right\},\\
  b(r_o') &= \int_0^{\theta_{\text{max}}}d\theta\ \sqrt{\cos\theta}\sin^2\theta J_1(k_b r_o'\sin\theta f_o/f_t),\\%\ \text{exp}\left\{ik_bz_o[1-(1/2)(f_o/f_t)^2\sin^2\theta]\right\},\\
  c(r_o') &= \int_0^{\theta_{\text{max}}}d\theta\ \sqrt{\cos\theta}\sin\theta(1-\cos\theta)J_2(k_b r_o'\sin\theta f_o/f_t).%\ \text{exp}\left\{ik_bz_o[1-(1/2)(f_o/f_t)^2\sin^2\theta]\right\}.
\end{align}
We can identify Eq. \ref{eq:elec} as the vector-valued \textit{coherent spread
  function} (CSF) of the microscope. The more familiar scalar-valued
CSF---sometimes called the amplitude transfer function---is only applicable to
cases where electromagnetic optics plays no role in the microscope---not true
for samples that contain dipole emitters.

To build our intuition about the CSF, we will rewrite the matrix multiplication
in Eq. \ref{eq:elec} in terms of the spherical harmonics. Notice that the $l=1$
spherical harmonics are the Cartesian components of the unit dipole axis (see
Eq. \ref{eq:harmonics}), so the CSF becomes
\begin{align}
  \tilde{\mb{e}}'_d(\ro{}', \so{}) &\propto
  \begin{bmatrix}
    [a(r_o') + c(r_o')\cos(2\phi_o')]y_1^{1}(\so{}) + c(r_o')\sin(2\phi_o')y_1^{-1}(\so{}) + 2ib(r_o')\cos\phi_o'y_1^{0}(\so{})\\
    c(r_o')\sin(2\phi_o')y_1^{1}(\so{}) + [a(r_o') - c(r_o')\cos(2\phi_o')]y_1^{-1}(\so{}) + 2ib(r_o')\sin\phi_o'y_1^{0}(\so{})\\
    0\\
  \end{bmatrix}.
\end{align}
By applying the paraxial approximation ($\sin\theta\approx\theta$ and
$\cos\theta\approx 1$), the integrals $a(r_o'), b(r_o')$ and $c(r_o')$ can be
evaluated in terms of Bessel functions. We can evaluate $a(r_o')$ and $b(r_o')$ with the
help of $\int_0^zdx\ xJ_0(ax) = zJ_1(az)/a$ and
$\int_0^zdx\ x^2J_1(ax) = z^2J_2(az)/a$, respectively, and $c(r_o') = 0$ because
$J_2(x)$ is zero to first order. In this case, the CSF simplifies to
\begin{align}
  \tilde{\mb{e}}'^{(p)}_d(\ro{}', \so{}) &\propto
  \begin{bmatrix}
    a^{(p)}(r_o')y_1^{1}(\so{}) + 2ib^{(p)}(r_o')\cos\phi_o'y_1^{0}(\so{})\\
    a^{(p)}(r_o')y_1^{-1}(\so{}) + 2ib^{(p)}(r_o')\sin\phi_o'y_1^{0}(\so{})\\
    0\\
  \end{bmatrix}\label{eq:paracsf},
\end{align}
where the integrals evaluate to
\begin{align}
  {a^{(p)}}(r_o') = \frac{J_1(2\pi \nu_or_o')}{\pi \nu_or_o'}, 
  &\hspace{2em}
    {b^{(p)}}(r_o') = \frac{\text{NA}}{n_o}\left[\frac{J_2(2\pi \nu_or_o')}{\pi \nu_or_o'}\right],  \label{eq:abparadef}
  \intertext{and we have substituted}
  \nu_o \equiv \frac{\text{NA}}{\lambda},&\hspace{2em}
  \text{NA} \equiv n_o\sin\theta_{\text{max}}.
\end{align}
Under the paraxial approximation the electric fields on the detector created by
a single dipole are composed of two parts, a parallel part and a radial part.
The coefficients of the $y_1^0$ spherical harmonic in Eq. \ref{eq:paracsf}
create the radial part of the field---the $z$ component of the dipole generates
a radial field on the detector with respect to the image point. The coefficients
of the $y_1^1$ and $y_1^{-1}$ spherical harmonics create a parallel field in the
detector plane parallel to the $x$ and $y$ components of the dipole. As the NA
increases and the paraxial approximation no longer applies, we begin to see the
$x$ and $y$ components of the dipole contribute to the $y$ and $x$ transverse
fields, respectively.

We can find the intensity in the detector plane due to a single dipole
by taking the squared modulus of the CSF
\begin{align}
  h'(\ro{}', \so{}) = \left|\tilde{\mb{e}}'_d(\ro{}', \so{})\right|^2 \label{eq:kernel}.
\end{align}
For convenience we keep the CSF written in terms of the spherical harmonics and
use a table of spherical harmonic products (see Appendix \ref{realspherical}) to
calculate the intensity in the detector plane as
\begin{equation}
  \begin{split}
  h'(\ro{}', \so{}) \propto &\left(a^2(r_o') + 2b^2(r_o') + c^2(r_o')\right)y_0^0(\so{}) -\frac{2\sqrt{15}}{5}a(r_o')c(r_o')\sin(2\phi_o')y_2^{-2}(\so{})\\ &+ \frac{1}{\sqrt{5}}\left(-a^2(r_o') + 4b^2(r_o') - c^2(r_o')\right)y_2^{0}(\so{}) +\frac{2\sqrt{15}}{5}a(r_o')c(r_o')\cos(2\phi_o')y_2^{2}(\so{}). \label{eq:kerna}
\end{split}
\end{equation}
We can identify Eq. \ref{eq:kerna} as the spatio-angular \textit{point spread
  function} (PSF) of the microscope. Writing the spatio-angular PSF in terms of
spherical harmonic functions has two advantages. First, it allows us to express
the spatio-angular PSF very concisely. Instead of considering the point spread
function for every possible dipole orientation, we only need to consider four
spatio-angular PSFs---one for each spherical harmonic. Second, the spherical
harmonic functions form an orthonormal basis for functions on the sphere---a
convenient fact that we will use later.

@@

The spatio-angular PSF under the paraxial approximation is given by
\begin{align}
      h'^{(p)}(\ro{}', \so{}) \propto \left({a^{(p)}}^2(r_o') + 2{b^{(p)}}^2(r_o')\right)y_0^0(\so{}) + \frac{1}{\sqrt{5}}\left(-{a^{(p)}}^2(r_o') + {4b^{(p)}}^2(r_o')\right)y_2^{0}(\so{})\label{eq:para}.
\end{align}
We rewrite the PSF in the following form to make plotting and interpretation easier
\begin{align}
  h'^{(p)}(\ro{}', \so{}) &= {h_0^0}^{(p)}(\rd{}')y_0^0(\so{}) + {h_2^0}^{(p)}(\rd{}')y_2^0(\so{}), \label{eq:concisepsf}\\
  {{h'}_0^0}^{(p)}(\ro{}') &\equiv {a^{(p)}}^2(r_o') + 2{b^{(p)}}^2(r_o'),\\
  {{h'}_2^0}^{(p)}(\ro{}') &\equiv \frac{1}{\sqrt{5}}\left[- {a^{(p)}}^2(r_o') + 4{b^{(p)}}^2(r_o')\right]. 
\end{align}

First, consider the coefficient on the $y_0^0$ spherical harmonic in Eq.
\ref{eq:para}. This coefficient is the point spread function for an angularly
uniform distribution of fluorophores. The first term of the coefficient
${a^{(p)}}^2(r_o')$ is the familiar Airy disk that arises from the contribution
of dipoles oriented in the transverse plane, while the second term
${b^{(p)}}^2(r_o')$ is a smaller factor that arises from dipoles oriented
outside of the transverse plane. This leads to an interesting conclusion---a
uniform distribution of dipoles has a point spread function that is slightly
wider than an Airy disk even in the paraxial approximation. The Airy disk is
usually derived using paraxial scalar optics while here we have used paraxial
electromagnetic optics. Therefore, we can consider the second term to be an
electromagnetic correction to the Airy disk. We will quantify this difference in
the next section.

Next, consider the coefficient on the $y_2^0$ spherical harmonic in
Eq. \ref{eq:para}. This coefficient is the spatial PSF for a distribution of
fluorophores proportional to $3\cos^2\vartheta_o - 1$. Counterintuitively, this
fluorophore distribution cannot exist because it would require a negative number
of fluorophores along some orientations; but if this distribution could exist,
then this coefficient would be its spatial PSF. Considering negative
distributions of fluorophores in our calculations should not be cause for
concern. The spherical harmonics span the space of functions on the sphere, so
any positive fluorophore distribution can be represented by the spherical
harmonics, and we never need to consider negative fluorophores.

@@

We can find the electric field on the detector created by this object by
multiplying $\bs{\mu}(\ro{}, \so{})$ with the CSF then integrating over all
positions and orientations in the object
\begin{align}
  \mb{e}'_d(\rd{}')  \propto \int_{\mathbb{S}^2}d\so{}\int_{\mathbb{R}^2}d\ro{} \ \tilde{\mb{e}}_d'(\rd{}' - \mb{r}_o, \so{})\bs{\mu}(\ro{}, \so{}).\label{eq:eforward}
\end{align}
Note that these integrals represent a vector sum---the coherence of the electric
fields radiated by the object can cause cancellations of the fields created on
the detector.
